Devito for large scale elastic modelling and anisotropic inversion

### 2020-04-21 paper.tex
Drop hyperlinks to Devito repo

It is with these specifications in mind that
\href{https://github.com/devitocodes/devito}{Devito}, a symbolic domain
specific language (DSL) and compiler for the automatic generation of
finite-difference stencils, has been developed. Originally deigned to
accelerate research and development in exploration geophysics, the high-level
interface, previously described in detail in \cite{devito-api}, is built on top
of \texttt{SymPy} \cite{sympy} and is inspired by the underlying mathematical
formulations and other DSLs such as FEniCS \cite{fenics} and Firedrake
\cite{firedrake}. This interface allows the user to formulate wave-equations,
and more generally time-dependent PDEs in a simple and mathematically coherent
way. The Devito compiler then automatically generates finite-difference
stencils from these mathematical expressions. One of the main advantages of
\href{https://github.com/devitocodes/devito}{Devito} over other
finite-difference DSLs is that generic expressions such as sparse operations
(i.e.~point source injection or localized measurements) are fully supported and
expressible in a high-level fashion. The second component of
\href{https://github.com/devitocodes/devito}{Devito} is its compiler (c.f
\cite{devito-compiler}) that generates highly optimized C code. The generated
code is then compiled at runtime for the hardware at hand.

Previous work focused on the DSL and compiler to highlight the potential
application and use cases of Devito. Here, we present a series of
extensions and applications to large-scale three-dimensional problem
sizes as encountered in exploration geophysics. These experiments are
carried out in Cloud-based HPC systems and include elastic forward
modelling using distributed-memory parallelism and imaging based on the
tilted transverse isotropic (TTI) wave-equation
(\cite{virieux, thomsen1986, zhang-tti, duveneck, louboutin2018segeow}).
These proof of concepts highlight two critical features: first, the
ability of the symbolic interface and the compiler to translate to
large-scale adjoint-based inversion problems that require massive
compute (since thousands of PDEs are solved) as well as large amounts of
memory (since the adjoint state method requires the forward model to be
saved in memory). Secondly, through the elastic modelling example, we
demonstrate that \href{https://github.com/devitocodes/devito}{Devito}
now fully supports and automates vectorial and second order tensorial
staggered-grid finite-differences with the same high-level API
previously presented for scalar fields defined on cartesian grids.

This paper is organized as follows: first, we provide a brief overview
of \href{https://github.com/devitocodes/devito}{Devito} and its symbolic
API and present the distributed memory implementation that allows
large-scale modeling and inversion by means of domain decomposition. We
then provide a brief comparison with a state of the art hand-coded wave
propagator to validate the performance previously benchmarked with the
roofline model
(\cite{patterson, devito-compiler, devito-api, louboutin2016ppf}).
Before concluding, results from the Cloud-based experiments discussed
above are presented, highlighting the vectorial and tensorial
capabilities of Devito.

@@

It is up to the \href{https://github.com/devitocodes/devito}{Devito}
compiler to translate the symbolic specification into C code. The
lowering of the input language down to C consists of several compilation
passes, some of which introducing performance optimizations that are the
key to fast code. Next to classic stencil optimizations (e.g., cache
blocking, alignment, SIMD and OpenMP parallelism),
\href{https://github.com/devitocodes/devito}{Devito} applies a series of
FLOP-reducing transformations as well as aggressive loop fusion. For a
complete treatment, the interested reader should refer to
\cite{devito-compiler}.

@@

In this section we illustrate the
\href{https://github.com/devitocodes/devito}{Devito} language showing
how to implement the acoustic wave-equation in isotropic media
%
\begin{equation}
\begin{cases}
 m \frac{d^2 u(t, x)}{dt^2} - \Delta u(t, x) = \delta(xs) q(t) \\
 u(0, .) = \frac{d u(t, x)}{dt}(0, .) = 0 \\
 d(t, .) = u(t, xr).
 \end{cases}
\label{acou}
\end{equation}
%
 The core of the \href{https://github.com/devitocodes/devito}{Devito}
symbolic API consists of three classes:

@@

The source term is handled separately from the PDE as a pointwise
operation called injection, while the measurement is handled
with an interpolation. By default,
\href{https://github.com/devitocodes/devito}{Devito} initializes all
\texttt{Function} data to 0, and thus automatically satisfies the zero
Dirichlet condition at \texttt{t=0}. The isotropic acoustic
wave-equation can then be implemented in
\href{https://github.com/devitocodes/devito}{Devito} as below

@@

The output of the \href{https://github.com/devitocodes/devito}{Devito}
compiler for the running example used in this section is available at
\href{https://github.com/mloubout/SC20Paper/tree/master/gencode}{CodeSample}
in \texttt{acou-so8.c}.

@@

Devito implements distributed-memory parallelism on top of MPI. The
design is such that users can almost entirely abstract away from it and
reuse non-distributed code as is. Given any
\href{https://github.com/devitocodes/devito}{Devito} code, just running
it as

@@

The domain decomposition occurs in Python upon creation of a
\texttt{Grid} object. Exploiting the MPI Cartesian topology abstraction,
\href{https://github.com/devitocodes/devito}{Devito} logically splits a
grid based on the number of available MPI processes (users are given an
``escape hatch'' to override Devito's default decomposition strategy).
\texttt{Function} and \texttt{TimeFunction} objects inherit the
\texttt{Grid} decomposition. For \texttt{SparseFunction} objects the
approach is different. Since a \texttt{SparseFunction} represents a
sparse set of points,
\href{https://github.com/devitocodes/devito}{Devito} looks at the
physical coordinates of each point and, based on the \texttt{Grid}
decomposition, schedules the logical ownership to an MPI rank. If a
sparse point lies along the boundary of two or more MPI ranks, then it
is duplicated in each of these ranks to be accessible by all neighboring
processes. Eventually, a duplicated point may be redundantly computed by
multiple processes, but any redundant increments will be discarded.

When accessing or manipulating data in a
\href{https://github.com/devitocodes/devito}{Devito} code, users have
the illusion to be working with classic NumPy arrays, while underneath
they are actually distributed. All manner of NumPy indexing schemes
(basic, slicing, etc.) are supported. In the implementation, proper
global-to-local and local-to-global index conversion routines are used
to propagate a read/write access to the impacted subset of ranks. For
example, consider the array

@@

Finally, we remark that while providing abstractions for distributed
data manipulation, \href{https://github.com/devitocodes/devito}{Devito}
does not support natively any mechanisms for parallel I/O. The
distributed NumPy arrays, however, provide a generic and flexible
infrastructure for the implementation of parallel I/O for any sort of
file format (e.g., see \cite{witte2018alf}).

@@

The main complexity of the TTI wave-equation is that the rotation of the
symmetry axis of the physics leads to rotated second-order finite-difference
stencils. In order to ensure numerical stability, these rotated
finite-difference operators are designed to be self-adjoint (c.f.
\cite{zhang2011, duveneck}). For example, we define the rotated second order
derivative with respect to $x$ as:
%
\begin{equation}
\begin{aligned}
  G_{\bar{x}\bar{x}} &= D_{\bar{x}}^T D_{\bar{x}} \\
  D_{\bar{x}} &= \cos(\mathbf{\theta})\cos(\mathbf{\phi})\frac{\mathrm{d}}{\mathrm{d}x} + \cos(\mathbf{\theta})\sin(\mathbf{\phi})\frac{\mathrm{d}}{\mathrm{d}y} - \sin(\mathbf{\theta})\frac{\mathrm{d}}{\mathrm{d}z}.
\end{aligned}
\label{rot}
\end{equation}
%
 We enable the simple expression of these complicated stencils in
\href{https://github.com/devitocodes/devito}{Devito} as
finite-difference shortcuts such as \texttt{u.dx} where \texttt{u}
is a \texttt{Function}. Such shortcuts are enabled not only for the
basic types but for generic composite expressions, for example
\texttt{(u\phantom{\ }+\phantom{\ }v.dx).dy}. As a consequence, the
rotated derivative defined in~\ref{rot} is implemented with
\href{https://github.com/devitocodes/devito}{Devito} in two lines as:

@@

Owing to the very high number of floating-point operations (FLOPs) needed per
grid point for the weighted rotated Laplacian, this anisotropic wave-equation
is extremely challenging to implement. As we show in Table~\ref{ttiFLOPs}, and
previously analysed in \cite{louboutin2016ppf}, the computational cost with
high-order finite-difference is in the order of thousands of FLOPs per grid
point without optimizations. The version without FLOP-reducing optimizations is
a direct translation of the discretized operators into stencil expressions (see
\texttt{tti-so8-unoptimized.c}). The version with optimizations employs
transformations such as common sub-expressions elimination, factorization, and
cross-iteration redundancy elimination -- the latter being key in removing
redundancies introduced by mixed derivatives. Implementing all of these
techniques manually is inherently difficult and laborious. Further, to obtain
the desired performance improvements it is necessary to orchestrate them with
aggressive loop fusion (for data locality), tiling (for data locality and
tensor temporaries), and potentially ad-hoc vectorization strategies (if
rotating registers are used). While an explanation of the optimization
strategies employed by \href{https://github.com/devitocodes/devito}{Devito} is
beyond the scope of this paper (see \cite{devito-compiler} for details), what
is emphasized here is that users can easily take full advantage of these
optimization without needed to concern themselves with the details.

@@

It is evident that developing an appropriate solver for the TTI wave-equation,
an endeavor involving complicated physics, mathematics, and engineering, is
exceptionally time-consuming and can lead to thousands of lines of code even for
a single choice of discretization. Verification of the results is no less
complicated, any minor error is effectively untrackable and any change to the
finite-difference scheme or to the time-stepping algorithm is difficult to
achieve without substantial re-coding. Another complication stems from the fact
that practitioners of seismic inversion are often geoscientists, not computer
scientists/programmers. Low level implementations from non-specialists can often
often lead to poorly performing code. However, if research codes are passed to
specialists in the domain of low level code optimization they often lack the
necessary geophysical domain knowledge, resulting in code that may lack a key
feature required by the geoscientist. Neither situation is conducive to
addressing the complexities that come with implementing codes based on the
latest geophysical insights in tandem with those from high-performance
computing. With \href{https://github.com/devitocodes/devito}{Devito} on the
other hand, both the forward and adjoint equations can be implemented in a few
lines of Python code as illustrated with the rotated operator in \#rotxpy. The
low level optimization element of the development is then taken care of under
the hood by the \href{https://github.com/devitocodes/devito}{Devito} compiler.

@@

We now demonstrate the scalability of
\href{https://github.com/devitocodes/devito}{Devito} to real-world applications
by imaging an industry-scale three-dimensional TTI subsurface model. This
imaging was carried out in the Cloud on Azure and takes advantage of recent work
to port conventional cluster code to the Cloud using a serverless approach. The
serverless implementation is detailed in \cite{witte2019TPDedas,
witte2019SEGedw} where the steps to run computationally and financially
efficient HPC workloads in the Cloud are described. This imaging project, in
collaboration with Azure, demonstrates the scalability and robustness of
\href{https://github.com/devitocodes/devito}{Devito} to large scale
wave-equation based inverse problems in combination with a cost-effective
serverless implementation of seismic imaging in the Cloud. In this example, we
imaged a synthetic three-dimensional anisotropic subsurface model that mimics a
realistic industry size problem with a realistic representation of the physics
(TTI). The physical size of the problem is \texttt{10kmx10kmx2.8km} discretized
on a \texttt{12.5m} grid with absorbing layers of width 40 grid points on each
side leading to \texttt{881x881x371} computational grid points ($\approx300$
Million grid points). The final image is the sum of 1500 single-source images:
100 single-source images were computed in parallel on the 200 nodes available
using two nodes per source experiment.

@@

The execution time is computed through Python-level timers prefixed by an MPI
barrier. The floating-point operations are counted once all of the symbolic
FLOP-reducing transformations have been performed during compilation.
\href{https://github.com/devitocodes/devito}{Devito} uses an in-house estimate
of cost, rather than \texttt{SymPy}'s estimate, to take care of some low-level
intricacies. For example, Devito's estimate ignores the cost of integer
arithmetic used for offset indexing into multi-dimensional arrays. To calculate
the total number of FLOPs performed,
\href{https://github.com/devitocodes/devito}{Devito} multiplies the
floating-point operations calculated at compile time by the size of the
iteration space, and it does that at the granularity of individual expressions.
Thanks to aggressive code motion, the amount of innermost-loop-invariant
sub-expressions in a \href{https://github.com/devitocodes/devito}{Devito}
Operator is typically negligible and therefore the
\href{https://github.com/devitocodes/devito}{Devito} estimate does not suffer
from this issue, or at least not, to the best of our knowledge, in a tangible
way. The Devito-reported GFLOP/s were also checked against those produced by
Intel Advisor on several single-node experiments: the differences -- typically
\href{https://github.com/devitocodes/devito}{Devito} underestimating the
achieved performance -- were always at most in the order of units, and therefore
negligible.

@@

While the subsurface image obtained in section \ref{d-imaging-example-on-azure}
utilized anisotropic propagators capable of mimicking intricate physics, in order
to model both the wave kinematics and amplitudes correctly, elastic
propagators are required. These propagators are, for example, extremely
important in global seismology since shear surface waves (which are
ignored in acoustic models) are the most hazardous. In this section, we exploit
the tensor algebra language introduced in
\href{https://github.com/devitocodes/devito}{Devito} v4.0 to express
an elastic model with compact and elegant notation.

@@

We have augmented the
\href{https://github.com/devitocodes/devito}{Devito} language with
tensorial objects to enable straightforward and
mathematically rigorous definitions of high-dimensional PDEs, such as
the elastic wave-equation defined in \eqref{elas1}. This implementation was
inspired by \cite{ufl}, a functional language for finite element
methods.

The extended \href{https://github.com/devitocodes/devito}{Devito}
language introduces two new types, \texttt{VectorFunction}/
\texttt{VectorTimeFunction} for vectorial objects such as the particle
velocity, and \texttt{TensorFunction}/\texttt{TensorTimeFunction}
for second-order tensor objects (matrices) such as the stress. These new
objects are constructed in the same manner as scalar
\texttt{Function} objects. They also automatically implement staggered
grid and staggered finite-differences with the possibility of half-node
averaging. Each component of a tensorial object -- a (scalar)
\href{https://github.com/devitocodes/devito}{Devito} \texttt{Function}
-- is accessible via conventional vector notation (i.e.
\texttt{v{[}0{]},\phantom{\ }t{[}0,1{]},....}).

@@

The \texttt{SymPy} expressions created by these commands can be
displayed with \texttt{sympy.pprint} as shown in
Figure~\ref{PrettyElas}. This representation reflects perfectly the
mathematics while maintaining computational portability and
efficiency through the
\href{https://github.com/devitocodes/devito}{Devito} compiler. The
complete generated code for the elastic wave-equation with and without
MPI is made available at
\href{https://github.com/mloubout/SC20Paper/tree/master/gencode}{CodeSample}
in \texttt{elastic-so8.c} and \texttt{elastic-so8-mpi.c}.

\begin{figure*}
\centering
\includegraphics[width=1.000\hsize]{./Figures/vel_symb.png}
\caption{Particle velocity update stencil. The stress update
is omitted for readability (the
equation does not fit into a single page). However, it can be found in
the \href{https://github.com/devitocodes/devito}{Devito} tutorial on
elastic modelling on github.}\label{PrettyElas}
\end{figure*}

@@

To demonstrate the efficacy of the elastic implementation outlined
above we utilized a broadly recognized 2D synthetic model, the elastic
Marmousi-ii\cite{versteeg927, marmouelas} model. The wavefields are
shown on Figure~\ref{ElasWf} and its corresponding elastic shot records
are displayed in Figure~\ref{ElasShot}. These two figures show that the
wavefield is, as expected, purely acoustic in the water layer
(\texttt{t\_\{xy\}=0}) and transitions correctly at the ocean bottom to
an elastic wavefield. We can also clearly see the shear wave-front
in the subsurface (at a depth of approximately 1km). Figures \ref{ElasWf}
and \ref{ElasShot} demonstrate that this high-level
\href{https://github.com/devitocodes/devito}{Devito} implementation
of the elastic wave-equation is effective and accurate.
Importantly, in constructing this model within the
\href{https://github.com/devitocodes/devito}{Devito} DSL framework,
computational technicalities such as the staggered grid analysis are
abstracted away. We note that the shot records displayed in Figure~\ref{ElasShot}
match the original data generated by the creator of this elastic model
(\cite{marmouelas}).

@@

Finally, three dimensional elastic data was modelled in the Cloud to
demonstrate the scalability of
\href{https://github.com/devitocodes/devito}{Devito} to cluster-size
problems. The model used in this experiment mimics a reference model in
geophysics known as the SEAM model \cite{fehler2011seam}, a
three dimensional extreme-scale synthetic representation of a
subsurface. The physical dimensions of the model are
\texttt{45kmx35kmx15km} discretized with a grid spacing of
\texttt{20mx20mx10m} leading to a computational grid of
\texttt{2250x1750x1500} grid points (5.9 billion grid points). One of
the main challenges of elastic modelling is the extreme memory cost owing
to the number of wavefields (a minimum of 21 fields in a three dimensional
propagator) that need to be stored:

@@

Because earlier performance benchmarks focus mainly on roofline model
analysis, we also compared the runtime of
\href{https://github.com/devitocodes/devito}{Devito} with a reference
open source hand-coded propagator in collaboration with its author for
completeness. This propagator, described in \cite{thorbecke} is a
state of the art elastic kernel (Equation~\ref{elas1}). For a fair
comparison, we ensure, in collaboration with the author of
\href{https://github.com/JanThorbecke/OpenSource.git}{fdelmodc} that the
physical and computational settings were identical. The setting were the
following:

@@

The runtime obtained for this problem for both propagators were
identical with less than one percent of difference. This similar runtime
were obtained both with the Intel compiler and the GNU compiler and we
ran the experiment with a fourth order and a sixth order discretization.
The kernels were executed five time each to ensure consistency between
the results and we consistently obtained similar runtimes. This
comparison illustrates the performance achieved with
\href{https://github.com/devitocodes/devito}{Devito} is at least on par
with hand-coded propagators. Considering we do not take adantage of the
\href{https://github.com/devitocodes/devito}{Devito} compiler full
capabilities in the two dimensional case, we are confident that the
generated code will be at least on par with hand-coded version for three
dimensional probalems and this comparison will be part of future work.

@@

The transition from academic size problems, such as the two-dimensional
acoustic wave-equation, to real-world application can be challenging in
particular when attempted as an afterthought. In this work, we showed
that thanks to design principles that aimed at this type of application
from the beginning, \href{https://github.com/devitocodes/devito}{Devito}
provides the scalability necessary. First, we showed that we provide a
high-level interface not only for simple scalar equations but for
coupled PDEs and allow the expression of non-trivial differential
operators in a simple and concise way. Second and most importantly, we
demonstrated that the compiler enables large-scale with state-of-the art
computational performance and programming paradigm. The single-node
performance is on par with hand-coded codes while providing the
flexibility from the symbolic interface, and multi-node parallelism is
integrated in the compiler and interface in a conservative way. Finally,
we demonstrated that our abstraction provides to performance portability
that enables both on-premise HPC and Cloud HPC.

>>>>

It is with these specifications in mind that Devito, a symbolic domain
specific language (DSL) and compiler for the automatic generation of
finite-difference stencils, has been developed. Originally deigned to
accelerate research and development in exploration geophysics, the
high-level interface, previously described in detail in \cite{devito-api},
is built on top of \texttt{SymPy} \cite{sympy} and is inspired by the
underlying mathematical formulations and other DSLs such as FEniCS
\cite{fenics} and Firedrake \cite{firedrake}. This interface allows the
user to formulate wave-equations, and more generally time-dependent PDEs
in a simple and mathematically coherent way. The Devito compiler then
automatically generates finite-difference stencils from these mathematical
expressions. One of the main advantages of Devito over other
finite-difference DSLs is that generic expressions such as sparse
operations (i.e.~point source injection or localized measurements) are
fully supported and expressible in a high-level fashion. The second
component of Devito is its compiler (c.f \cite{devito-compiler}) that
generates highly optimized C code. The generated code is then compiled at
runtime for the hardware at hand.

Previous work focused on the DSL and compiler to highlight the potential
application and use cases of Devito. Here, we present a series of
extensions and applications to large-scale three-dimensional problem sizes
as encountered in exploration geophysics. These experiments are carried
out in Cloud-based HPC systems and include elastic forward modelling using
distributed-memory parallelism and imaging based on the tilted transverse
isotropic (TTI) wave-equation (\cite{virieux, thomsen1986, zhang-tti,
duveneck, louboutin2018segeow}).  These proof of concepts highlight two
critical features: first, the ability of the symbolic interface and the
compiler to translate to large-scale adjoint-based inversion problems that
require massive compute (since thousands of PDEs are solved) as well as
large amounts of memory (since the adjoint state method requires the
forward model to be saved in memory). Secondly, through the elastic
modelling example, we demonstrate that Devito now fully supports and
automates vectorial and second order tensorial staggered-grid
finite-differences with the same high-level API previously presented for
scalar fields defined on cartesian grids.

This paper is organized as follows: first, we provide a brief overview of
Devito and its symbolic API and present the distributed memory
implementation that allows large-scale modeling and inversion by means of
domain decomposition. We then provide a brief comparison with a state of
the art hand-coded wave propagator to validate the performance previously
benchmarked with the roofline model (\cite{patterson, devito-compiler,
devito-api, louboutin2016ppf}).  Before concluding, results from the
Cloud-based experiments discussed above are presented, highlighting the
vectorial and tensorial capabilities of Devito.

@@

It is up to the Devito compiler to translate the symbolic specification
into C code. The lowering of the input language down to C consists of
several compilation passes, some of which introducing performance
optimizations that are the key to fast code. Next to classic stencil
optimizations (e.g., cache blocking, alignment, SIMD and OpenMP
parallelism), Devito applies a series of FLOP-reducing transformations as
well as aggressive loop fusion. For a complete treatment, the interested
reader should refer to \cite{devito-compiler}.

@@

In this section we illustrate the Devito language showing how to implement
the acoustic wave-equation in isotropic media
%
\begin{equation}
\begin{cases}
 m \frac{d^2 u(t, x)}{dt^2} - \Delta u(t, x) = \delta(xs) q(t) \\
 u(0, .) = \frac{d u(t, x)}{dt}(0, .) = 0 \\
 d(t, .) = u(t, xr).
 \end{cases}
\label{acou}
\end{equation}
%
 The core of the Devito
symbolic API consists of three classes:

@@

The source term is handled separately from the PDE as a pointwise
operation called injection, while the measurement is handled with an
interpolation. By default, Devito initializes all \texttt{Function} data
to 0, and thus automatically satisfies the zero Dirichlet condition at
\texttt{t=0}. The isotropic acoustic wave-equation can then be implemented
in Devito as below

@@

The output of the Devito compiler for the running example used in this
section is available at
\href{https://github.com/mloubout/SC20Paper/tree/master/gencode}{CodeSample}
in \texttt{acou-so8.c}.

@@

Devito implements distributed-memory parallelism on top of MPI. The design
is such that users can almost entirely abstract away from it and reuse
non-distributed code as is. Given any Devito code, just running it as

@@

The domain decomposition occurs in Python upon creation of a \texttt{Grid}
object. Exploiting the MPI Cartesian topology abstraction, Devito
logically splits a grid based on the number of available MPI processes
(users are given an ``escape hatch'' to override Devito's default
decomposition strategy).  \texttt{Function} and \texttt{TimeFunction}
objects inherit the \texttt{Grid} decomposition. For
\texttt{SparseFunction} objects the approach is different. Since a
\texttt{SparseFunction} represents a sparse set of points, Devito looks at
the physical coordinates of each point and, based on the \texttt{Grid}
decomposition, schedules the logical ownership to an MPI rank. If a sparse
point lies along the boundary of two or more MPI ranks, then it is
duplicated in each of these ranks to be accessible by all neighboring
processes. Eventually, a duplicated point may be redundantly computed by
multiple processes, but any redundant increments will be discarded.

When accessing or manipulating data in a Devito code, users have the
illusion to be working with classic NumPy arrays, while underneath they
are actually distributed. All manner of NumPy indexing schemes (basic,
slicing, etc.) are supported. In the implementation, proper
global-to-local and local-to-global index conversion routines are used to
propagate a read/write access to the impacted subset of ranks. For
example, consider the array

@@

Finally, we remark that while providing abstractions for distributed
data manipulation, Devito
does not support natively any mechanisms for parallel I/O. The
distributed NumPy arrays, however, provide a generic and flexible
infrastructure for the implementation of parallel I/O for any sort of
file format (e.g., see \cite{witte2018alf}).

@@

The main complexity of the TTI wave-equation is that the rotation of the
symmetry axis of the physics leads to rotated second-order finite-difference
stencils. In order to ensure numerical stability, these rotated
finite-difference operators are designed to be self-adjoint (c.f.
\cite{zhang2011, duveneck}). For example, we define the rotated second order
derivative with respect to $x$ as:
%
\begin{equation}
\begin{aligned}
  G_{\bar{x}\bar{x}} &= D_{\bar{x}}^T D_{\bar{x}} \\
  D_{\bar{x}} &= \cos(\mathbf{\theta})\cos(\mathbf{\phi})\frac{\mathrm{d}}{\mathrm{d}x} + \cos(\mathbf{\theta})\sin(\mathbf{\phi})\frac{\mathrm{d}}{\mathrm{d}y} - \sin(\mathbf{\theta})\frac{\mathrm{d}}{\mathrm{d}z}.
\end{aligned}
\label{rot}
\end{equation}
%

We enable the simple expression of these complicated stencils in Devito
as finite-difference shortcuts such as \texttt{u.dx} where \texttt{u} is
a \texttt{Function}. Such shortcuts are enabled not only for the basic
types but for generic composite expressions, for example
\texttt{(u\phantom{\ }+\phantom{\ }v.dx).dy}. As a consequence, the
rotated derivative defined in~\ref{rot} is implemented with Devito in two
lines as:

@@

Owing to the very high number of floating-point operations (FLOPs) needed per
grid point for the weighted rotated Laplacian, this anisotropic wave-equation
is extremely challenging to implement. As we show in Table~\ref{ttiFLOPs}, and
previously analysed in \cite{louboutin2016ppf}, the computational cost with
high-order finite-difference is in the order of thousands of FLOPs per grid
point without optimizations. The version without FLOP-reducing optimizations is
a direct translation of the discretized operators into stencil expressions (see
\texttt{tti-so8-unoptimized.c}). The version with optimizations employs
transformations such as common sub-expressions elimination, factorization, and
cross-iteration redundancy elimination -- the latter being key in removing
redundancies introduced by mixed derivatives. Implementing all of these
techniques manually is inherently difficult and laborious. Further, to obtain
the desired performance improvements it is necessary to orchestrate them with
aggressive loop fusion (for data locality), tiling (for data locality and
tensor temporaries), and potentially ad-hoc vectorization strategies (if
rotating registers are used). While an explanation of the optimization
strategies employed by Devito is beyond the scope of this paper (see
\cite{devito-compiler} for details), what is emphasized here is that users
can easily take full advantage of these optimization without needed to
concern themselves with the details.

@@

It is evident that developing an appropriate solver for the TTI wave-equation,
an endeavor involving complicated physics, mathematics, and engineering, is
exceptionally time-consuming and can lead to thousands of lines of code even for
a single choice of discretization. Verification of the results is no less
complicated, any minor error is effectively untrackable and any change to the
finite-difference scheme or to the time-stepping algorithm is difficult to
achieve without substantial re-coding. Another complication stems from the fact
that practitioners of seismic inversion are often geoscientists, not computer
scientists/programmers. Low level implementations from non-specialists can often
often lead to poorly performing code. However, if research codes are passed to
specialists in the domain of low level code optimization they often lack the
necessary geophysical domain knowledge, resulting in code that may lack a key
feature required by the geoscientist. Neither situation is conducive to
addressing the complexities that come with implementing codes based on the
latest geophysical insights in tandem with those from high-performance
computing. With Devito on the other hand, both the forward and adjoint
equations can be implemented in a few lines of Python code as illustrated
with the rotated operator in \#rotxpy. The low level optimization element
of the development is then taken care of under the hood by the Devito
compiler.

@@

We now demonstrate the scalability of Devito to real-world applications by
imaging an industry-scale three-dimensional TTI subsurface model. This
imaging was carried out in the Cloud on Azure and takes advantage of
recent work to port conventional cluster code to the Cloud using a
serverless approach. The serverless implementation is detailed in
\cite{witte2019TPDedas, witte2019SEGedw} where the steps to run
computationally and financially efficient HPC workloads in the Cloud are
described. This imaging project, in collaboration with Azure, demonstrates
the scalability and robustness of Devito to large scale wave-equation
based inverse problems in combination with a cost-effective serverless
implementation of seismic imaging in the Cloud. In this example, we imaged
a synthetic three-dimensional anisotropic subsurface model that mimics a
realistic industry size problem with a realistic representation of the
physics (TTI). The physical size of the problem is
\texttt{10kmx10kmx2.8km} discretized on a \texttt{12.5m} grid with
absorbing layers of width 40 grid points on each side leading to
\texttt{881x881x371} computational grid points ($\approx300$ Million grid
points). The final image is the sum of 1500 single-source images: 100
single-source images were computed in parallel on the 200 nodes available
using two nodes per source experiment.

@@

The execution time is computed through Python-level timers prefixed by an
MPI barrier. The floating-point operations are counted once all of the
symbolic FLOP-reducing transformations have been performed during
compilation.  Devito uses an in-house estimate of cost, rather than
\texttt{SymPy}'s estimate, to take care of some low-level intricacies. For
example, Devito's estimate ignores the cost of integer arithmetic used for
offset indexing into multi-dimensional arrays. To calculate the total
number of FLOPs performed, Devito multiplies the floating-point operations
calculated at compile time by the size of the iteration space, and it does
that at the granularity of individual expressions.  Thanks to aggressive
code motion, the amount of innermost-loop-invariant sub-expressions in an
\texttt{Operator} is typically negligible and therefore the Devito estimate
does not suffer from this issue, or at least not, to the best of our
knowledge, in a tangible way. The Devito-reported GFLOP/s were also
checked against those produced by Intel Advisor on several single-node
experiments: the differences -- typically Devito underestimating the
achieved performance -- were always at most in the order of units, and
therefore negligible.

@@

While the subsurface image obtained in section \ref{d-imaging-example-on-azure}
utilized anisotropic propagators capable of mimicking intricate physics, in order
to model both the wave kinematics and amplitudes correctly, elastic
propagators are required. These propagators are, for example, extremely
important in global seismology since shear surface waves (which are
ignored in acoustic models) are the most hazardous. In this section, we
exploit the tensor algebra language introduced in Devito v4.0 to express
an elastic model with compact and elegant notation.

@@

We have augmented the Devito language with tensorial objects to enable
straightforward and mathematically rigorous definitions of
high-dimensional PDEs, such as the elastic wave-equation defined in
\eqref{elas1}. This implementation was inspired by \cite{ufl}, a
functional language for finite element methods.

The extended Devito language introduces two new types,
\texttt{VectorFunction}/ \texttt{VectorTimeFunction} for vectorial objects
such as the particle velocity, and
\texttt{TensorFunction}/\texttt{TensorTimeFunction} for second-order
tensor objects (matrices) such as the stress. These new objects are
constructed in the same manner as scalar \texttt{Function} objects. They
also automatically implement staggered grid and staggered
finite-differences with the possibility of half-node averaging. Each
component of a tensorial object -- a (scalar) Devito \texttt{Function} --
is accessible via conventional vector notation (i.e.
\texttt{v{[}0{]},\phantom{\ }t{[}0,1{]},....}).

@@

The \texttt{SymPy} expressions created by these commands can be displayed
with \texttt{sympy.pprint} as shown in Figure~\ref{PrettyElas}. This
representation reflects perfectly the mathematics while maintaining
computational portability and efficiency through the Devito compiler. The
complete generated code for the elastic wave-equation with and without MPI
is made available at
\href{https://github.com/mloubout/SC20Paper/tree/master/gencode}{CodeSample}
in \texttt{elastic-so8.c} and \texttt{elastic-so8-mpi.c}.

\begin{figure*}
\centering
\includegraphics[width=1.000\hsize]{./Figures/vel_symb.png}
\caption{Particle velocity update stencil. The stress update
is omitted for readability (the equation does not fit into a single page).
However, it can be found in the Devito tutorial on elastic modelling
on GitHub.}\label{PrettyElas}
\end{figure*}

@@

To demonstrate the efficacy of the elastic implementation outlined
above we utilized a broadly recognized 2D synthetic model, the elastic
Marmousi-ii\cite{versteeg927, marmouelas} model. The wavefields are
shown on Figure~\ref{ElasWf} and its corresponding elastic shot records
are displayed in Figure~\ref{ElasShot}. These two figures show that the
wavefield is, as expected, purely acoustic in the water layer
(\texttt{t\_\{xy\}=0}) and transitions correctly at the ocean bottom to
an elastic wavefield. We can also clearly see the shear wave-front
in the subsurface (at a depth of approximately 1km). Figures \ref{ElasWf}
and \ref{ElasShot} demonstrate that this high-level Devito implementation
of the elastic wave-equation is effective and accurate.  Importantly, in
constructing this model within the Devito DSL framework, computational
technicalities such as the staggered grid analysis are abstracted away. We
note that the shot records displayed in Figure~\ref{ElasShot} match the
original data generated by the creator of this elastic model
(\cite{marmouelas}).

@@

Finally, three dimensional elastic data was modelled in the Cloud to
demonstrate the scalability of Devito to cluster-size problems. The model
used in this experiment mimics a reference model in geophysics known as
the SEAM model \cite{fehler2011seam}, a three dimensional extreme-scale
synthetic representation of a subsurface. The physical dimensions of the
model are \texttt{45kmx35kmx15km} discretized with a grid spacing of
\texttt{20mx20mx10m} leading to a computational grid of
\texttt{2250x1750x1500} grid points (5.9 billion grid points). One of the
main challenges of elastic modelling is the extreme memory cost owing to
the number of wavefields (a minimum of 21 fields in a three dimensional
propagator) that need to be stored:

@@

Because earlier performance benchmarks focus mainly on roofline model
analysis, we also compared the runtime of Devito with a reference
open source hand-coded propagator in collaboration with its author for
completeness. This propagator, described in \cite{thorbecke} is a
state of the art elastic kernel (Equation~\ref{elas1}). For a fair
comparison, we ensure, in collaboration with the author of
\href{https://github.com/JanThorbecke/OpenSource.git}{fdelmodc} that the
physical and computational settings were identical. The setting were the
following:

@@

The runtime obtained for this problem for both propagators were identical
with less than one percent of difference. This similar runtime were
obtained both with the Intel compiler and the GNU compiler and we ran the
experiment with a fourth order and a sixth order discretization.  The
kernels were executed five time each to ensure consistency between the
results and we consistently obtained similar runtimes. This comparison
illustrates the performance achieved with Devito is at least on par with
hand-coded propagators. Considering we do not take adantage of the Devito
compiler full capabilities in the two dimensional case, we are confident
that the generated code will be at least on par with hand-coded version
for three dimensional probalems and this comparison will be part of future
work.

@@

The transition from academic size problems, such as the two-dimensional
acoustic wave-equation, to real-world application can be challenging in
particular when attempted as an afterthought. In this work, we showed that
thanks to design principles that aimed at this type of application from
the beginning, Devito provides the scalability necessary. First, we showed
that we provide a high-level interface not only for simple scalar
equations but for coupled PDEs and allow the expression of non-trivial
differential operators in a simple and concise way. Second and most
importantly, we demonstrated that the compiler enables large-scale with
state-of-the art computational performance and programming paradigm. The
single-node performance is on par with hand-coded codes while providing
the flexibility from the symbolic interface, and multi-node parallelism is
integrated in the compiler and interface in a conservative way. Finally,
we demonstrated that our abstraction provides to performance portability
that enables both on-premise HPC and Cloud HPC.
